---
title: "GSEA analysis by cluster"
author: "José Manuel Gómez Silva"
date: "`r Sys.Date()`"
output: # html_notebook
---

```{r setup, include=FALSE}
all_times <- list() # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = TRUE,
  echo = TRUE,
  engine.opts = list(bash = "-l")
)
```


```{r libraries, include = FALSE}
library(Seurat)
library(ggplot2)
library(future) # Allows parallelization in Seurat.

# Set up Seurat pararell computing.
options(parallelly.fork.enable = TRUE)
plan("multicore", workers = parallel::detectCores()) # Adjust threads for memory use.

# Import user's libraries.
source(file = "~/Documents/SC_Prostate/Scripts/SC_utils/scGSEAmarkers.R")
```


# GSEA analysis as helper for cluster manuel annotation.

The purpose of this script is to obtain a GSEA enrichment analysis for all the positive markers of each of the clusters that are expressed at least on 50% of the cells.

## 1. Import data and set the identities.

```{r import_raw_data}

# SCP_data <- readRDS("~/Documents/SC_Prostate/Data/SC_Prostate_processed.rds")
SCP_data <- SetIdent(SCP_data, value = SCP_data$RNA_snn_res.0.1)
```

## 2. Plot clusters.

```{r cluster_annotation_vis, fig.height=6, fig.width=10}

DimPlot(SCP_data, reduction = "umap", label = TRUE, repel = TRUE, label.size = 4) + theme(axis.title = element_text(size = 15), legend.text = element_text(size = 10), axis.text = element_text(size = 10)) + guides(colour = guide_legend(override.aes = list(size = 3)))
```

## 3. Get foldchange.

```{r}

# Join the per-run splitted layers.
SCP_data <- JoinLayers(SCP_data, assay = "RNA")
```


We use the FindAllMarkers function to easily obtain the foldchange for the desired genes. The use of the wilcoxon test has no impact, is just for optimizing computation time.

We stablish a threshold for features that are expressed in at least 50% of the cells (a marker should be widely expressed).

We base our analysis on the C8 dataset of GSEA's MSigDB containing cell identity markers obtained from SC experiments.

```{r}

cluster_markers <- FindAllMarkers(object = SCP_data,
                                  test.use = "wilcox", # Use wilcox as DE method as it is well-optimized with presto.
                                  assay = "RNA",
                                  slot = "data", # Use log-transformed CPMs.
                                  # We need all genes for the enrichment analysis:
                                  only.pos = FALSE,
                                  min.pct = 0.5, # We keep the pct threshold to avoid underrepresented genes.
                                  logfc.threshold = 0,
                                  verbose = TRUE
)

# Load the pathways database from GSEA.
reference_markers <- fgsea::gmtPathways(gmt.file = "~/Documents/SC_Prostate/Data/cell_markers_GSEA.gmt")
```
We now compute the GSEA analysis using both positive and negative markers to increase statistical power but the output will only contain cell identities based on positive markers (entries with a positive Normalized Enrichment Score or NES) and we will keep only results with a p-adj below 1e-6.

```{r}

result_list <- scGSEAmarkers(cluster_markers, reference_markers, padj.threshold = 1e-6, only.pos = TRUE, workers = parallel::detectCores())
```
Export each cluster information as a single data.frame:

```{r}

path <- "~/Documents/SC_Prostate/Output/GSEA_markers_res=0.1/"

for (cluster in names(result_list)) {
  df <- result_list[[cluster]]
  df <- apply(df,2,as.character) # Transform each column to character to avoid conflicts with lists.
  filename <- paste0(path, cluster, ".csv")
  write.csv(df, file = filename, row.names = FALSE)
  cat("Exported dataframe", cluster, "to", filename, "\n")
}

```

Session Info.

```{r session_info}

sessionInfo()
RNGkind()
```