---
title: "Resolution testing."
author: "José Manuel Gómez Silva"
date: "`r Sys.Date()`"
output: # html_notebook
---

```{r setup, include=FALSE}

all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = TRUE,
  echo = TRUE,
  engine.opts = list(bash = "-l")
)
```

# Resolution value fine-tuning.

As reported by the Seurat team. A resolution value between 0.4 - 1.2 works well for an assay of \~3k cells.

Here we implement a loop that tests different resolution values with an increment of 0.1 between these two values.

First, we import all necessary libraries.

```{r libraries, include = FALSE}

library(Seurat)
library(dplyr)
library(patchwork)
library(sctransform)
library(ggplot2)
library(future) # Allows parallelization in Seurat.

# Set up Seurat pararell computing.
options(parallelly.fork.enable = TRUE)
plan("multicore", workers = parallel::detectCores())
```

Load the data and apply the basic pipeline in one line.

SCTransform:

```{r SCTransform}

resolutions = seq(from = 0.1, to = 2, by = 0.1)
  
test <- readRDS("~/Documents/SC_Prostate/Data/SC_Prostate_raw.rds") %>% 
  PercentageFeatureSet(pattern = "^MT-", col.name = 'percent.mt') %>%
  split(f = .$orig.ident) %>%
  SCTransform(vst.flavor = "v2", vars.to.regress = 'percent.mt', verbose = FALSE) %>% 
  RunPCA(assay = "SCT", verbose = FALSE) %>% 
  IntegrateLayers(method = HarmonyIntegration, normalization.method = "SCT", verbose = FALSE) %>% 
  FindNeighbors(dims = 1:30, reduction = "harmony", verbose = FALSE) %>% 
  FindClusters(resolution = resolutions, verbose = FALSE) %>% 
  RunUMAP(dims = 1:30, reduction = "harmony", verbose = FALSE)
```

Standard normalization:

```{r standard_norm}

resolutions = seq(from = 0.1, to = 2, by = 0.1)
  
test <- readRDS("~/Documents/SC_Prostate/Data/SC_Prostate_raw.rds") %>% 
  PercentageFeatureSet(pattern = "^MT-", col.name = 'percent.mt') %>%
  split(f = .$orig.ident) %>%
  NormalizeData(scale.factor = 1e6, assay = "RNA", verbose = FALSE) %>% 
  FindVariableFeatures(nfeatures = 3000, assay = "RNA", verbose = FALSE) %>%
  ScaleData(assay = "RNA", vars.to.regress = "percent.mt", verbose = FALSE) %>%
  RunPCA(assay = "RNA", verbose = FALSE) %>% 
  IntegrateLayers(method = HarmonyIntegration, verbose = FALSE) %>% 
  FindNeighbors(dims = 1:20, reduction = "harmony", verbose = FALSE) %>% 
  FindClusters(resolution = resolutions, verbose = FALSE) %>% 
  RunUMAP(dims = 1:20, reduction = "harmony", verbose = FALSE)
```


In the script below we created a loop that test all values for resolution in a certain range. It also plots the resulting umaps and storage both the Seurat Objects and plots generated in two different lists (output_list and plot_list). We tested the resolution in a range between 0.4 - 1.2 as the Seurat teams indicates that is suitable for a dataset of up to 3k cells.

```{r}

clusterMetadata <- grep("_snn_res.\\d+.?\\d*", colnames(test[[]]), value = TRUE)

# Create empty lists to storage data.
resolution_values <- list()
nClusters <- list()

for (resValue in clusterMetadata) {
  
  resolution <- as.numeric(stringr::str_extract(resValue, "\\d+.?\\d*"))
  
  Idents(test) <- tibble::deframe(test[[resValue]])
  
  # Plot each group of clusters generated, save the images and store the plots in a list.
  plot <- DimPlot(test, reduction = 'umap', label = TRUE, repel = TRUE, label.size = 4) + ggtitle(paste0("Resolution = ", resolution)) + theme(axis.title = element_text(size = 15), legend.text = element_text(size = 10), axis.text = element_text(size = 10)) + guides(colour = guide_legend(override.aes = list(size = 3)))
  
  ggsave(filename = paste0("~/Documents/SC_Prostate/Output/resolution_fine-tuning/resolution=", resolution, ".jpg"), height = 6, width = 10, plot = plot, quality = 50)
  
  # To visualize the number of clusters obtain for each resolution value, we plot number of clusters generated vs the resolution value.
  
  # Storage resolution values in a list.
  resolution_values[[length(resolution_values) + 1]] <- resolution
  
  # Storage number of clusters in a lists. 
  # This access the number of clusters in the SeuratObject, which is storage as a factor inside the active.ident slot. We obtain the levels, transform them into integers, extract the maximum value and sum 1 (Seurat clusters numeration starts by 0).
  output_clusters <- max(as.integer(levels(test@active.ident))) + 1
  nClusters[[length(nClusters) + 1]] <- output_clusters
  
  cat("Calculated and plotted for resolution = ", resolution, "\n")
  
}
```

```{r plotting}

# Plot number of clusters against resolution.
plot(resolution_values, nClusters)
```

Session Info.

```{r session_info}

sessionInfo()
RNGkind()
```