---
title: "Resolution testing."
author: "José Manuel Gómez Silva"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}

all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = TRUE,
  echo = TRUE,
  engine.opts = list(bash = "-l")
)
```

# Resolution value fine-tuning.

As reported by the Seurat team. A resolution value between 0.4 - 1.2 works well for an assay of 3k cells.

Here we implement a loop that tests different resolution values with an increment of 0.1 between these two values.

We will use Edu's pre-processed data.

First, we import all necessary libraries.

```{r libraries, include = FALSE}

library(Seurat)
library(dplyr)
library(patchwork)
library(sctransform)
library(ggplot2)
library(celldex) # Cell annotation.
library(SingleR) # Cell annotation.
library(parallel)  # detectCores()
library(future) # Allows parallelization in Seurat.

# Set up Seurat pararell computing.
options(parallelly.fork.enable = TRUE)
plan("multicore", workers = detectCores())
```

Load the data and apply the basic pipeline in one line.

```{r previous_steps}

test <- readRDS("../Data/run1/C1_Seurat_Edu.rds") %>% PercentageFeatureSet(pattern = "^MT-", col.name = 'percent.mt') %>% SCTransform(vst.flavor = "v2", vars.to.regress = 'percent.mt') %>% RunPCA(assay = "SCT") %>% FindNeighbors(dims = 1:30)


```

In the script below we created a loop that test all values for resolution in a certain range. It also plots the resulting umaps and storage both the Seurat Objects and plots generated in two different lists (output_list and plot_list). We tested the resolution in a range between 0.4 - 1.2 as the Seurat teams indicates that is suitable for a dataset of up to 3k cells.

```{r}

# Set the range of values with an increment of 0.1
start_value <- 0.1
end_value <- 1.2
increment <- 0.1


# Create a loop to test the function with incremental values.

# Create emtpy lists to storage data.
output_list <- list()
plot_list <- list()
resolution_values <- list()
nClusters <- list()

# This equivalent of range() in python.
for (i in seq(start_value, end_value, by = increment)) {
  
  # Generate a new SeuratObjecto for each resolution value and storage it on a list.
  output <- FindClusters(test, resolution = i, verbose = FALSE) %>% RunUMAP(dims = 1:30, verbose = FALSE)
  output_list[[length(output_list) + 1]] <- output
  
  # Plot each group of clusters generated, save the images and store the plots in a list.
  plot <- DimPlot(output, reduction = 'umap', label = TRUE, repel = TRUE, label.size = 4) + ggtitle(paste0("Resolution = ", i)) + theme(axis.title = element_text(size = 15), legend.text = element_text(size = 10), axis.text = element_text(size = 10)) + guides(colour = guide_legend(override.aes = list(size = 3)))
  plot_list[[length(plot_list) + 1]] <- plot
  
  ggsave(filename = paste0("../Output/images/resolution=", i, ".jpg"), height = 6, width = 10, plot = plot, quality = 50)
  
  # To visualize the number of clusters obtain for each resolution value, we plot number of clusters generated vs the resolution value.
  
  # Storage resolution values in a list.
  resolution_values[[length(resolution_values) + 1]] <- i
  
  # Storage number of clusters in a lists. 
  # This access the number of clusters in the SeuratObject, which is storage as a factor inside a df in the object. We access the df as a column, accesing the first row. Whe obtain the levels, transform them into integers, extract the maximum value and sum 1 (seurat clusters numeration starts by 0).
  output_clusters <- max(as.integer(levels(output[["seurat_clusters"]][1,]))) + 1
  nClusters[[length(nClusters) + 1]] <- output_clusters
  
  cat("Calculated and plotted for resolution = ", i, "\n")
}

# Clean variables to save memory.
rm(output)
rm(output_clusters)
rm(plot)
rm(i)

# Transform the list to vectors in order to plot them.
resolution_values <- unlist(resolution_values)
nClusters <- unlist(nClusters)

# Plot number of clusters against resolution.
plot(resolution_values, nClusters, ylim = c(0,max(nClusters)))
```

After analyzing the results, we will use a resolution value of 0.8, as the region between 0.6-0.9 seems the most stable, and 0.8 offers the maximum number of distinct clusters within that range while also being the default for the FindClusters() function.
