---
title: "SC_Prostate analysis"
author: "José Manuel Gómez Silva"
date: "`r Sys.Date()`"
output: # html_notebook
---

```{r setup, include=FALSE}

all_times <- list() # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  error = TRUE,
  echo = TRUE,
  engine.opts = list(bash = "-l")
)
```

```{r libraries, include = TRUE}

library(Seurat)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(celldex) # Cell annotation.
library(SingleR) # Cell annotation.
library(future) # Allows parallelization in Seurat.

# Set up Seurat pararell computing.
options(parallelly.fork.enable = TRUE)
plan("multicore", workers = parallel::detectCores())

# Import user's libraries.
source(file = "~/Documents/SC_Prostate/Scripts/SC_utils/FeatureScatterGradient.R")
```

# 1. Load filtered run data.

First, we search trough the project directory for each filtered raw data and load each SeuratObject.

```{r get_runs}

# List the data directory.
run_dirs <- list.dirs("../Data",recursive = FALSE)

# Filter only the run folders.
# Grepl stands for "logical grep" so it only outputs TRUE or FALSE. It accepts an array of strings as an input, search for teh given pattern and returns another array with TRUE or FALSE if the pattern matches with the string in that position of the array.
run_dirs <- run_dirs[grepl("run",run_dirs)]

# lapply stands for "list apply", and allows to apply a certain function over all elements of an array: lapply(X, FUN). You can apply an R-defined function or create you own as shown belox.
rds_objects <- lapply(X = run_dirs, FUN = function(x) {

  rds_file <- list.files(path = x, pattern = "*_filtered.rds",full.names = T) # For each run directory list the filtered RDS object.

  # If there is something in the directory list...
    if (length(rds_file) > 0){
      rds_object <- readRDS(rds_file) # Load the RDS object in to a list.
      rds_object@meta.data$orig.ident <- basename(x) # Add the original path in the metadata to keep track of data.
      return(rds_object) # Return the list of RDS object and...

    } else {
      return(NULL) # If the directory list is empty return null.
    }
  }
)

# sapply is similar to lapply but returns a vector of the same lenght instead of the same element as the input (i.e. lapply returns a list is input is a list, but sapply always returns a vector).
rds_objects[sapply(rds_objects, is.null)] <- NULL # Filter emtpy objects.
```

# 2. Merge Seurat Objects.

We merge the 4 datasets into a unique Seurat Object. We confirm the merge by checking the total number off cells.

```{r merge_runs, fig.height=6, fig.width=10}

# Reduce() allows to reduce a list of values into a single value by applying a binary operation over then iteratively. That is, takes the first two elements of a list, x and y, apply a the function (fun(x,y)) creating the reduced value a, then takes the third element of the list, z, and apply the function to c and z: fun(c, z) and so on till the whole list is reduce to one single value. In this case, we apply Seruat's merge to merge all objects sequentially into one single SeuratObject for the whole data set.
# add.cell.ids = add the project name to each specific run to its cells.
SCP_data <- Reduce(function(x,y) merge(x,y, add.cell.ids = c(x@project.name,y@project.name)), rds_objects)

# check that the size is correct
for (rds_object in rds_objects){
  print(dim(rds_object))
}

print(dim(SCP_data))
```

Removal of big variables used for data merging:

```{r remove_objects, inlude=FALSE}

# remove run objects
rm(rds_object, rds_objects, run_dirs)
```

Saving raw merged data:

```{r save_raw_data, include=FALSE}

saveRDS(SCP_data, file = "~/Documents/SC_Prostate/Data/SC_Prostate_raw.rds")
```

# 3. Visualization of QC metrics.

```{r load_raw_data, include = FALSE}

# SCP_data <- readRDS(file = "~/Documents/SC_Prostate/Data/SC_Prostate_raw.rds")
```


Checking the QC metrics of the combine dataset.

```{r QC_visualization, fig.height=6, fig.width=10}

# Visualize QC metrics as a violin plot
VlnPlot(SCP_data, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# Zoom in on nFeature_RNA violin plot.
VlnPlot(SCP_data, features = "nFeature_RNA", ncol = 1) + ylim(0, 2500) + NoLegend()

# Zoom in on nCount_RNA violin plot.
VlnPlot(SCP_data, features = "nCount_RNA", ncol = 1) + ylim(0, 25000) + NoLegend()

# Visualize relationships in metadata to detect outliers with FeatureScatter function
plot1 <- FeatureScatter(SCP_data, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatterGradient(SCP_data, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", gradient = "percent.mt", upper.limit = 100, scale.colors = "viridis")
plot1 + plot2
```

Visualizing metrics by sample:

```{r sample_QC, fig.height=6, fig.width=10}

# Visualize QC metrics as a violin plot
VlnPlot(SCP_data, features = "nFeature_RNA", group.by = "Sample_Name") + NoLegend()
VlnPlot(SCP_data, features = "nCount_RNA", group.by = "Sample_Name") + NoLegend()
VlnPlot(SCP_data, features = "percent.mt", group.by = "Sample_Name") + NoLegend()
```

Apparently, run2 and run4 have higher minimum counts than run1 and run3. This could led to batch effect that will be corrected in later steps. Despite this, samples and run appear to have a pretty similar distribution of QC metrics.

## 3.1. Removal of low cell count samples.

If we check the cell number per samples, some of them have specially low cell counts. To avoid this low-informative samples, we establish a threshold of at least 100 cels for a sample to be considered for the analyis.

```{r calculate_cells_sample}
cells_per_sample <- SCP_data@meta.data %>% group_by(Sample_Name) %>% summarise(Cells = n())
cells_per_sample
```

```{r plot_cells_sample, fig.height=6, fig.width=10}

ggplot(cells_per_sample, aes(x=Sample_Name, y=Cells, fill = Sample_Name)) +
  geom_bar(stat = "identity", color = "black") +
  geom_hline(aes(yintercept=100), linetype="dashed", color="black") +
  theme_classic() + 
  theme(legend.position="none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)
        )
  
```
```{r filter_samples}

filteredSamples <- cells_per_sample %>% filter(Cells>=100) %>% select(Sample_Name) %>% unlist(.)
SCP_data <- subset(SCP_data, Sample_Name %in% filteredSamples)
SCP_data
```

# 4. Normalize and remove batch effect.

Batch effect is considered the presence of certain unique characteristics within a specific dataset that make it differentiable from the rest when merging all the data. As we are merging data from different runs, variations in the samples processing and further cell manipulation and sequencing could induce those batch effect.

Before the merge, we annotate the cell with its run of origin, so we can regress out that variable in the normalization process to minimize batch effect.

Seurat offers methods to integrate data from different batches (different sequencing runs, different samples, etc.). In v5, a new method called IntegrateLayers they easily perform the whole pipeline in a few steps. The first of them is to split the data according to the source of batch effect (in our case by run).

```{r split_data}

# The split method splits the given slot according to the indicated information.
SCP_data[["RNA"]] <- split(SCP_data[["RNA"]], f = SCP_data$orig.ident)
```

# 4.1. Normalize and apply data integration.

It is necessary to normalize the data before integration. According to maintainers (https://github.com/satijalab/seurat/issues/4811, https://github.com/satijalab/seurat/issues/7585) it doesn't mind to normalize before or after splitting.

We now run normalization, PCA and the integration.

```{r batch_effect, fig.height=6, fig.width=10}

# Standard normalization.
SCP_data <- NormalizeData(SCP_data, scale.factor = 1e6, assay = "RNA") # scale.factor = 1e6 means we are using CPMs.
SCP_data <- FindVariableFeatures(SCP_data, nfeatures = 3000, assay = "RNA")
SCP_data <- ScaleData(SCP_data, assay = "RNA", vars.to.regress = "percent.mt") # regress out % mito.

# Run PCA.
SCP_data <- RunPCA(SCP_data, assay = "RNA")

# Integration.
SCP_data <- IntegrateLayers(
  object = SCP_data, method = HarmonyIntegration,
  verbose = FALSE
)
```

There are several methods for integration, but according to some sources (https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1850-9), the best ones are Harmony, Seurat's CCA and scVII (requires setting up a conda environment and installing reticulate, see documentation). We choose Harmony for being well-integrated with Seruat, performing pretty well with low computational effort, and better respecting the data original structure when compared against Seurat's CCA.

If we visualize the PCA plot, we appreciate a bigger overlap between runs.

```{r fig.height=6, fig.width=10}

Idents(SCP_data) <- "orig.ident"
DimPlot(SCP_data, reduction = "pca")
```

## 4.2. Set dimensionality.

While the elbow appears around the 15th PC, we extend the included PCs up to the 20th following the guidelines by the Seurat developers.

```{r elbwol_plot, fig.height=6, fig.width=10}

ElbowPlot(SCP_data, ndims = 50) # choose 20
```

The heatmap and the feature plot give us information about the significance and composition of each PC.

```{r PCA_heatmap, fig.height=15, fig.width=10}

DimHeatmap(SCP_data, dims = 1:20, balanced = TRUE)
```

```{r PCA_top_features, fig.height=30, fig.width=15}

VizDimLoadings(SCP_data, dims = 1:20, reduction = "pca")
```


# 5. Clustering.

Clustering resolution fine-tuning is performed by a separated .Rmd file (Resolution fine-tuning.Rmd), giving 1 as an appropriate value for clustering.

It is important to notice that a new reduction is created called "harmony". We need to indicate this as the reduction to use in the clustering.

```{r clustering}

resolutions = c(0.2, 0.4, 1)

# Resolution fine-tuned in a separate Rmd.
SCP_data <- FindNeighbors(SCP_data, dims = 1:20, reduction = "harmony")
SCP_data <- FindClusters(SCP_data, resolution =  resolutions) # Default resolution = 0.8
SCP_data <- RunUMAP(SCP_data, dims = 1:20, reduction = "harmony")
```


## 5.1. Cluster QC.

Cluster QC shows homogeneous characteristics across the clusters with the exception of cluster 22 (abnormal low number of features). Despite what it might look like, this actually conforms and advantage, as if that specific group of cells give inconsistent results, because they have been isolated in their own cluster we can move forward just ignoring or removing it.

```{r cluster_QC, fig.height=6, fig.width=10}

# Visualize QC metrics as a violin plot
VlnPlot(SCP_data, features = "nFeature_RNA") + NoLegend()
VlnPlot(SCP_data, features = "nCount_RNA") + NoLegend()
VlnPlot(SCP_data, features = "percent.mt") + NoLegend()

# Zoom in on nFeature_RNA violin plot.
VlnPlot(SCP_data, features = "nFeature_RNA", ncol = 1) + ylim(0, 2500) + NoLegend()

# Visualize relationships in metadata to detect outliers with FeatureScatter function
plot1 <- FeatureScatter(SCP_data, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(SCP_data, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + NoLegend() + plot2 + NoLegend()
```

# 6. Annotation.

## 6.1. By-cell annotation.

Cell-by-cell annotation identifies 30 different cell types based on the HumanPrimaryCellAtlas, which distribution pretty much coincides with the cluster layout.

```{r annotation}

# Using celldex and SingleR packages.
# Download reference data from celldex.
reference <- HumanPrimaryCellAtlasData()

# Join layers before conversion. SingleR uses the RNA assay.
merged.SCexp <- JoinLayers(SCP_data, assay = "RNA") 

# Convert Seurat object into a SingleCellExperiment Object for SingleR input.
merged.SCexp <- as.SingleCellExperiment(merged.SCexp)

SingleR.annotation <- SingleR(test = merged.SCexp, ref = reference, assay.type.test = "logcounts", labels = reference$label.main, num.threads = parallel::detectCores())

SCP_data[["cell.labels"]] <- SingleR.annotation$labels
```

```{r annotation_vis_cell, fig.height=6, fig.width=10}

DimPlot(SCP_data, reduction = "umap", group.by = "cell.labels", label = TRUE, repel = TRUE, label.size = 4) + theme(axis.title = element_text(size = 15), legend.text = element_text(size = 10), axis.text = element_text(size = 10)) + guides(colour = guide_legend(override.aes = list(size = 3)))
```

## 6.2. By-cluster annotation.

We create our own composite cluster names by adding the cluster annotation to the cluster number provided by Suerat. Thus, what we see is an estimation of the cell type that better fits the genetic landscape of each cluster.

```{r cluster_annotation}

# Obtaining a vector containing the cluster of each cell in order.

# Get the factor contained in the SeuratObject with all this information.
clusters <- SCP_data@meta.data[["seurat_clusters"]]

# The cluster information for each cell is contain as a factor which levels coincide with the total number of clusters found by FindClusters(). An approach to transform this factor into a character vector is the following:
# Obtain the list of clusters with levels(clusters). This outputs a character vector containing the levles of the factor. After that, we use the factor itself as an index to access the levels vector. When using a factor as an index, R does not use the labels itself (which in this case are string, so if used as indexes would cause an error), but the internal numeric index the factor contains. That way, for each cluster label associated with a cell in the factor, we access its numeric index and map it to the levels vectors (which coincides), thus obtaining each cell label as an unique character value. Each cell label is then storage as a character (the as.character is added as a control method since SingleR only admits strings as labels) in a vector. The vector contains the cluster label for each cell as a character value in the same order as each cell appears in the dataset, so the by-cluster annotation doesn't assign the cells to an incorrect cluster.
clusters <- as.character(levels(clusters)[clusters])

# reference <- HumanPrimaryCellAtlasData()

# # Join layers before conversion. SingleR uses the RNA assay.
# merged.SCexp <- JoinLayers(SCP_data, assay = "RNA") 
# merged.SCexp <- as.SingleCellExperiment(SCP_data)

# We input the cluster vector using the clusters parameter.
SingleR.annotation <- SingleR(test = merged.SCexp, ref = reference, assay.type.test = "logcounts", labels = reference$label.main, clusters = clusters, num.threads = parallel::detectCores())

SCP_data[["cluster.labels"]] <- SingleR.annotation$labels
```

```{r set_ids}

# We composite the cluster name. That way when 2 clusters' names are the same Seurat doesn't merge the labels.

# Get clusters levels accessing the SeuratObject variable as a df and then accessing the df as a column.
cluster_number <- levels(SCP_data[["seurat_clusters"]][1, ])

# Get annotation labels.
cluster_annotation <- SingleR.annotation$labels

# Since cluster levels and labels are in the same order, we composite the new names using paste0 (sort of equivalent to fstrings in python).
new.clusters.ids <- paste0(cluster_number, "-", cluster_annotation)

# Add names to each value of the clusters id vector so Seurat can take it as a valid input for RenameIdents.
names(new.clusters.ids) <- levels(SCP_data)
SCP_data <- RenameIdents(SCP_data, new.clusters.ids)
SCP_data[["cell.cluster.labels"]] <- Idents(SCP_data)
```

```{r cluster_annotation_vis, fig.height=6, fig.width=10}

DimPlot(SCP_data, reduction = "umap", label = TRUE, repel = TRUE, label.size = 4) + theme(axis.title = element_text(size = 15), legend.text = element_text(size = 10), axis.text = element_text(size = 10)) + guides(colour = guide_legend(override.aes = list(size = 3)))
```

```{r save_processed_data, include=FALSE}

saveRDS(SCP_data, file = "~/Documents/SC_Prostate/Data/SC_Prostate_processed.rds")
```

Session Info.

```{r session_info}

sessionInfo()
RNGkind()
```
